### 2018-08-01 mlr_summary.tex
sum fixes

\section{Linear Regression}
One of the most basic algorithms for estimating a real valued function is linear regression.
Given a set of multivariate inputs $x_i \in \dom{R}^m$ and corresponding real valued outputs $y_i \in \dom{R}$ the following \enquote{loss} $L$ is to be minimized by a linear function $f_\beta:\dom{R}^m\to\dom{R}$.
\eqns{&
L(\beta) = \sum_{i=1}^n (\;y_i - f_\beta(x_i)\;)^2 
&&\\&
f_\beta(x) = x\T\beta \with\beta\in\dom{R}^m
&&}
This is the mean squared error and we want to minimize it using a linear function.
Lets call the matrix of transposed $x_i$'s $X$ and the vector of $y_i$'s $y$.
This allows for rewriting the loss as follows.
\eqnsnn{&
L(\beta) = (y-X\beta)\T(y-X\beta)
&&\\&
X = \mat{x_1\T\\\vdots\\x_n\T} \quad y = \mat{y_1\\\vdots\\y_n}
&&}
When minimizing we seek the roots of the derivative of the loss, by doing so we can derive the optimal solution for beta.
\eqnsnn{
\deriv{L(\beta)}{\beta} = 0 ~&\equivalent~ -(y-X\beta)\T X - (y-X\beta)\T X = -2*(y-X\beta)\T X = 0
&&\\&
\equivalent (y-X\beta)\T X = X\T y - X\T X\beta = 0
&&\\&
\equivalent X\T y = X\T X\beta
&&\\&
\equivalent (X\T X)^{-1}X\T y = \beta
&&
}
\subsection{Non-Linear Features}
Sadly the very simple linear regression algorithm only allows for linear functions to be estimated. 
Fortunately we can make things non-linear by defining a feature function which maps an input $x$ to feature space like so:
\eqnsnn{
\phi(x) \in \dom{R}^l \with l\in\dom{N}
&&}
For real valued $x\in\dom{R}$ we can create the quadratic feature vector using the following.
\eqnsnn{
\phi(x) = \mat{1&x&x^2}\T
&&}
When using linear regression on these features we end up with a function that is quadratic in $x$ but linear in features $\phi(x)$.
\eqnsnn{&
X = \mat{\phi(x_1)\T\\\vdots\\\phi(x_n)\T}
&&\\&
f_\beta(x) = \phi(x)\T\beta = \phi(x)_1*\beta_1 + \phi(x)_2*\beta_2 + \phi(x)_3*\beta_3 = \beta_1 + x\beta_2 + x^2\beta_3
&&
}
Even more sophisticated features could be using gaussian functions at the datapoints which then amounts to a radial basis function.
\eqnsnn{&
\phi(x) = \mat{1 & \phi_1(x) & \cdots & \phi_N(x)}\T \with \phi_i(x) = e^{-\frac{1}{2}||x_i-x||_2^2}
&&}

\subsection{Regularization}
Maybe you may have noticed that the linear regression algorithm may fail in case that $(X\T X)$ is not invertible.
To ensure invertibility a tiny quantity can be added to or subtracted from the diagonal of the matrix like so: $(X\T X - \lambda I)$.
Interestingly, the optimal solution to the following loss has this exact form.
\eqnsnn{
L(\beta) = (y&-X\beta)\T(y-X\beta) + \lambda*\beta\T\beta
&&\\
\deriv{L(\beta)}{\beta} = 0 &\equivalent -2(y-X\beta)\T X + 2\lambda*\beta\T = 0
&&\\
&\equivalent \lambda*\beta\T - (y-X\beta)\T X = 0
&&\\
&\equivalent \lambda*\beta - X\T y + X\T X\beta = 0 
&&\\
&\equivalent X\T y = X\T X\beta - \lambda*\beta = (X\T X - \lambda I) \beta
&&\\
&\equivalent (X\T X - \lambda I)^{-1}X\T y = \beta
&&} 
The loss function can be viewed as a combination of penalties for $\beta$.
The first penalty is the mean squared error from before which is large when $\beta$ cannot reproduce the samples $y$ well.
The second penalty is for $\beta$ of long lenghts, which means that short $\beta$ with small coefficients are preferred.
These two penalties probably are contradictive and the optimal beta will be somewhere in between the individual minima for the penalties.
This kind of regularization using $\lambda\beta\T\beta$ is called $L^2$ regularization and is a special case of the general Tikhonov regularization formulating constraints on $\beta$ as $\beta\T\Gamma\T\Gamma\beta$ with Tikhonov matrix $\Gamma$.
$\Gamma\beta$ is a linear mapping which can be chosen to also formulate special constraints on beta like differences between coefficients, or projections to different vector spaces.

@@

\subsection{Cross Validation}
In order to find the perfect regularization parameter $\lambda$ in order for our model to generalize well we need to try some out.
This can be done using cross validation.
In cross validation, we divide our data set into $K$ euqually sized parts (randomly), then we train our model using K-1 parts of the data and validate the model by computing the mean suared error using the missing part.
This is done k times so that each part has been used for validation once and the errors are averaged.
The following pseudo code should make the algorithm clear.
\begin{algorithmic}[1]
\State Split data $D$ into $K$ equally sized parts $\{D_1,\dots,D_K\}$ randomly
\For{$k=1\dots K$}
\State train on data $D_{\backslash D_k}$ yielding $\beta_i$
\State calculate MSE using $D_k$ ~ $err_k = L^{MSE}(\beta_k,D_k)$
\EndFor
\State calculate total MSE as $err=\frac{1}{K}\sum_{k=1}^K \frac{err_k}{|D_k|}$ 
\end{algorithmic}
The total error will be a measure for generalization of the model, using this measure an optimal $\lambda$ can be found by trying different $\lambda$ and adaptively tightening the search space.
For the start it may be useful to try lambdas of different orders of magnitude first.

@@

To get used to the plane stuff again, lets review the point-normal form of a plane.
In the simplest case where the plane intersects the origin, we can check if a point is in the plane by projecting its position vector $x$ onto the plane normal $n$. When $x$ is orthogonal to $n$ the projection $x\T n = 0$ and $x$ is in the nullspace of $n$ that is the plane.
When the plane does not intersect the origin we first need to translate the coordinatesystem so that it does.
We can do this by subtarcting a support vector of the plane $p_0$ (a position vector of a point that is part of the plane) from $x$, resulting in the check $(x-p_0)\T n = 0$ which can be rewritten as follows.
\eqnsnn{&
(x-p_0)\T n = 0 \equivalent x\T n + d = 0 \quad\text{with constant}\quad d = -p_0\T n
&&}
If $n$ is of unit length, then the distance of $x$ to the plane is given by the projection onto the normal $dist=x\T n + d$.
When not normalized then the distance is given by the following.
\eqnsnn{&
dist = (x-p_0)\T \frac{n}{|n|} = \frac{x\T n}{|n|} - \frac{p_0\T n}{|n|} = \frac{x\T n + d}{|n|}
&&}
This means that the projection on the normal is the distance to the plane scaled by the normals length $x\T n + d = dist*|n|$.
For a fixed projection value, this means, the smaller the normal the greater the distance.
In fact for a projection of value $1$ the distance is $\frac{1}{|n|}$.
We want the margin, that is the \enquote{width} of the separator, to be maximal.
If we require projections to be at least $1$ (or $-1$ depending on the side) the margin is $\frac{1}{2*|n|}$.

@@

So the projection $z_i = V_P\T(x_i-\mu)$ is the projecting the \enquote{centered} $x_i$.
For the derivation of $V_P$ lets assume that $x_i$ are already centered ($E(x)=0$).
What we want to minimize is the projection error, which we can estimate by back projecting the projection of $x_i$ and comparing it with the initial $x_i$.
\eqnsnn {&
\hat{V}_P = \argmin_{V_P} \sum_{i=1}^N ||x_i- V_PV_P\T x_i||^2 
&&}
This time we will derive the optimum by observing the properties of the equation instead of root finding in the derivative.
Lets reformulate the loss a bit.
\eqnsnn {
L(V_P) &= \sum_{i=1}^N (x_i- V_PV_P\T x_i)\T(x_i- V_PV_P\T x_i)
&&\\
&= \sum_{i=1}^N x_i\T x_i - 2 x_i\T V_P V_P\T x_i 
+ x_i\T V_P V_P\T V_P V_P\T x_i
&&\\
& = \sum_{i=1}^N x_i\T x_i - 2x_i\T V_P V_P\T x_i  + x_i\T V_P V_P\T x_i
&&\\
&= \sum_{i=1}^N x_i\T x_i - x_i\T V_P V_P\T x_i
&&\\
&= C - \sum_{i=1}^N x_i\T V_P V_P\T x_i \with C = \sum_{i=1}^N x_i\T x_i
&&\\
&= C - \sum_{i=1}^N || V_P\T x_i ||^2 \mathtext{($V_P\T$ composed of row vectors $v_j\T$)}
&&\\
&= C - \sum_{i=1}^N \sum_{j=1}^P (v_j\T x_i)*(v_j\T x_i)
 \;=\; C - \sum_{j=1}^P \sum_{i=1}^N  (x_i\T v_j)*(x_i\T v_j)
&&\\
&= C - \sum_{j=1}^P ||X v_j||^2 \mathtext{($X$ composed of row vectors $x_i\T$)}
&&\\
&= C - \sum_{j=1}^P v_j\T X\T X v_j
&&}
So whatwe can see from this, is that the loss will be minimal if we maximize $\sum_{j=1}^P v_j\T X\T X v_j$.
We could again find the root of its derivative but, it will be more fun with some clever factorization of $X\T X$, which is the Eigendecomposition. 
Since $X\T X$ is a symmetric matrix, its Eigendecomposition is
\eqnsnn{&
X\T X = Q\Lambda Q\T \mathtext{with orthonormal} Q \mathtext{and diagonal} \Lambda
&&\\
& \argmax_{v_{1:P}} \sum_{j=1}^P v_j\T Q\Lambda Q\T v_j
&&} 
Suppose the eigenvlus $\lambda_1,\dots,\lambda_M$ are sorted in value descending order on the diagonal of $\Lambda$.
To maximize the the above expression, we need to choose $v_{1:P}$ to be the first $P$ eigenvectors $q_i$ corresponding to the $P$ largest eigenvalues. 
If we could, we would choose all $v_j$ to be the first eigenvector, as this would truly yield the maximum, but due to our constraint of $v_j$s being orthonormal to each other we can't do that.

>>>>

\section{Linear Regression}
One of the most basic algorithms for estimating a real valued function is linear regression.
Given a set of multivariate inputs $x_i \in \dom{R}^m$ and corresponding real valued outputs $y_i \in \dom{R}$ the following \enquote{loss} $L$ is to be minimized by a linear function $f_\beta:\dom{R}^m\to\dom{R}$.
\eqns{&
L(\beta) = \sum_{i=1}^n (\;y_i - f_\beta(x_i)\;)^2 
&&\\&
f_\beta(x) = x\T\beta \with\beta\in\dom{R}^m
&&}
This is the mean squared error and we want to minimize it using a linear function.
Lets call the matrix of transposed $x_i$'s $X$ and the vector of $y_i$'s $y$.
This allows for rewriting the loss as follows.
\eqnsnn{&
L(\beta) = (y-X\beta)\T(y-X\beta)
&&\\&
X = \mat{x_1\T\\\vdots\\x_n\T} \quad y = \mat{y_1\\\vdots\\y_n}
&&}
When minimizing we seek the roots of the derivative of the loss, by doing so we can derive the optimal solution for beta.
\eqnsnn{
\deriv{L(\beta)}{\beta} = 0 ~&\equivalent -(y-X\beta)\T X - (y-X\beta)\T X = -2*(y-X\beta)\T X = 0
&&\\&
\equivalent (y-X\beta)\T X = X\T y - X\T X\beta = 0
&&\\&
\equivalent X\T y = X\T X\beta
&&\\&
\equivalent (X\T X)^{-1}X\T y = \beta
&&
}
\subsection{Non-Linear Features}
Sadly the very simple linear regression algorithm only allows for linear functions to be estimated. 
Fortunately we can make things non-linear by defining a feature function which maps an input $x$ to feature space like so:
\eqnsnn{
\phi(x) \in \dom{R}^l \with l\in\dom{N}
&&}
For real valued $x\in\dom{R}$ we can create the quadratic feature vector using the following.
\eqnsnn{
\phi(x) = \mat{1&x&x^2}\T
&&}
When using linear regression on these features we end up with a function that is quadratic in $x$ but linear in features $\phi(x)$.
\eqnsnn{&
X = \mat{\phi(x_1)\T\\\vdots\\\phi(x_n)\T}
&&\\&
f_\beta(x) = \phi(x)\T\beta = \phi(x)_1*\beta_1 + \phi(x)_2*\beta_2 + \phi(x)_3*\beta_3 = \beta_1 + x\beta_2 + x^2\beta_3
&&
}
Even more sophisticated features could be using gaussian functions at the datapoints which then amounts to a radial basis function.
\eqnsnn{&
\phi(x) = \mat{1 & \phi_1(x) & \cdots & \phi_N(x)}\T \with \phi_i(x) = e^{-\frac{1}{2}||x_i-x||_2^2}
&&}

\subsection{Regularization}
Maybe you may have noticed that the linear regression algorithm may fail in case that $(X\T X)$ is not invertible.
To ensure invertibility a tiny quantity can be added to the diagonal of the matrix like so: $(X\T X + \lambda I)$.
Interestingly, the optimal solution to the following loss has this exact form.
\eqnsnn{
L(\beta) = (y&-X\beta)\T(y-X\beta) + \lambda*\beta\T\beta
&&\\
\deriv{L(\beta)}{\beta} = 0 &\equivalent -2(y-X\beta)\T X + 2\lambda*\beta\T = 0
&&\\
&\equivalent \lambda*\beta\T - (y-X\beta)\T X = 0
&&\\
&\equivalent \lambda*\beta - X\T y + X\T X\beta = 0 
&&\\
&\equivalent X\T y = X\T X\beta + \lambda*\beta = (X\T X + \lambda I) \beta
&&\\
&\equivalent (X\T X + \lambda I)^{-1}X\T y = \beta
&&} 
The loss function can be viewed as a combination of penalties for $\beta$.
The first penalty is the mean squared error from before which is large when $\beta$ cannot reproduce the samples $y$ well.
The second penalty is for $\beta$ of long lenghts, which means that short $\beta$ with small coefficients are preferred.
These two penalties probably are contradictive and the optimal beta will be somewhere in between the individual minima for the penalties.
This kind of regularization using $\lambda\beta\T\beta$ is called $L^2$ regularization and is a special case of the general Tikhonov regularization formulating constraints on $\beta$ as $\beta\T\Gamma\T\Gamma\beta$ with Tikhonov matrix $\Gamma$.
$\Gamma\beta$ is a linear mapping which can be chosen to also formulate special constraints on beta like differences between coefficients, or projections to different vector spaces.

@@

\subsection{Cross Validation}
In order to find the perfect regularization parameter $\lambda$ for our model to generalize well, we need to try some $\lambda$s out.
This can be done using cross validation.
In cross validation, we divide our data set into $K$ euqually sized parts (randomly), then we train our model using K-1 parts of the data and validate the model by computing the mean suared error using the missing part.
This is done k times so that each part has been used for validation once and the errors are averaged.
The following pseudo code should make the algorithm clear.
\begin{algorithmic}[1]
\State Split data $D$ into $K$ equally sized parts $\{D_1,\dots,D_K\}$ randomly
\For{$k=1\dots K$}
\State train on data $D_{\backslash D_k}$ yielding $\beta_i$
\State calculate MSE using $D_k$ ~ $err_k = L^{MSE}(\beta_k,D_k)$
\EndFor
\State calculate total MSE as $err=\frac{1}{K}\sum_{k=1}^K \frac{err_k}{|D_k|}$ 
\end{algorithmic}
The total error will be a measure for generalization of the model, using this measure an optimal $\lambda$ can be found by trying different $\lambda$ and adaptively tightening the search space.
For the start it may be useful to try lambdas of different orders of magnitude first.

@@

To get used to the plane stuff again, lets review the point-normal form of a plane.
In the simplest case where the plane intersects the origin, we can check if a point is in the plane by projecting its position vector $x$ onto the plane normal $n$. When $x$ is orthogonal to $n$ the projection $x\T n = 0$ and $x$ is in the nullspace of $n$ that is the plane.
When the plane does not intersect the origin we first need to translate the coordinatesystem so that it does.
We can do this by subtarcting a support vector of the plane $p_0$ (a position vector of a point that is part of the plane) from $x$, resulting in the check $(x-p_0)\T n = 0$ which can be rewritten as follows.
\eqnsnn{&
(x-p_0)\T n = 0 \equivalent x\T n + d = 0 \quad\text{with constant}\quad d = -p_0\T n
&&}
If $n$ is of unit length, then the distance of $x$ to the plane is given by the projection onto the normal $dist=x\T n + d$.
When not normalized then the distance is given by the following.
\eqnsnn{&
dist = (x-p_0)\T \frac{n}{|n|} = \frac{x\T n}{|n|} - \frac{p_0\T n}{|n|} = \frac{x\T n + d}{|n|}
&&}
This means that the projection on the normal is the distance to the plane scaled by the normals length $x\T n + d = dist*|n|$.
For a fixed projection value, this means, the smaller the normal the greater the distance.
In fact for a projection of value $1$ the distance is $\frac{1}{|n|}$.
We want the margin, that is the \enquote{width} of the separator, to be maximal.
If we require projections to be at least $1$ (or $-1$ depending on the side) the margin is $\frac{2}{|n|}$.

@@

So the projection $z_i = V_P\T(x_i-\mu)$ is the projecting the \enquote{centered} $x_i$.
For the derivation of $V_P$ lets assume that $x_i$ are already centered ($E(x)=0$).
What we want to minimize is the projection error, which we can estimate by back projecting the projection of $x_i$ and comparing it with the initial $x_i$.
\eqnsnn {&
\hat{V}_P = \argmin_{V_P} \sum_{i=1}^N ||x_i- V_PV_P\T x_i||^2 
&&}
This time we will derive the optimum by observing the properties of the equation instead of root finding in the derivative.
Lets reformulate the loss a bit.
\eqnsnn {
L(V_P) &= \sum_{i=1}^N (x_i- V_PV_P\T x_i)\T(x_i- V_PV_P\T x_i)
&&\\
&= \sum_{i=1}^N x_i\T x_i - 2 x_i\T V_P V_P\T x_i 
+ x_i\T V_P V_P\T V_P V_P\T x_i
&&\\
& = \sum_{i=1}^N x_i\T x_i - 2x_i\T V_P V_P\T x_i  + x_i\T V_P V_P\T x_i
&&\\
&= \sum_{i=1}^N x_i\T x_i - x_i\T V_P V_P\T x_i
&&\\
&= C - \sum_{i=1}^N x_i\T V_P V_P\T x_i \with C = \sum_{i=1}^N x_i\T x_i
&&\\
&= C - \sum_{i=1}^N || V_P\T x_i ||^2 \mathtext{($V_P\T$ composed of row vectors $v_j\T$)}
&&\\
&= C - \sum_{i=1}^N \sum_{j=1}^P (v_j\T x_i)*(v_j\T x_i)
 \;=\; C - \sum_{j=1}^P \sum_{i=1}^N  (x_i\T v_j)*(x_i\T v_j)
&&\\
&= C - \sum_{j=1}^P ||X v_j||^2 \mathtext{($X$ composed of row vectors $x_i\T$)}
&&\\
&= C - \sum_{j=1}^P v_j\T X\T X v_j
&&}
So what we can see from this, is that the loss will be minimal if we maximize $\sum_{j=1}^P v_j\T X\T X v_j$.
We could again find the root of its derivative but, it will be more fun with some clever factorization of $X\T X$, which is the Eigendecomposition. 
Since $X\T X$ is a symmetric matrix, its Eigendecomposition is
\eqnsnn{&
X\T X = Q\Lambda Q\T \mathtext{with orthonormal} Q \mathtext{and diagonal} \Lambda
&&\\
& \argmax_{v_{1:P}} \sum_{j=1}^P v_j\T Q\Lambda Q\T v_j
&&} 
Suppose the eigenvalues $\lambda_1,\dots,\lambda_M$ are sorted in value descending order on the diagonal of $\Lambda$.
To maximize the above expression, we need to choose $v_{1:P}$ to be the first $P$ eigenvectors $q_i$ corresponding to the $P$ largest eigenvalues. 
If we could, we would choose all $v_j$ to be the first eigenvector, as this would truly yield the maximum, but due to our constraint of $v_j$s being orthonormal to each other we can't do that.
